dyngen: simulating developing single cells

### 2019-09-03 content.tex
update

We define the abundance levels of pre-mRNA, mRNA and protein of gene $G$ as $w_G$ $x_G$ and $y_G$ respectively. Five reactions affect the abundance levels of these molecules: transcription, splicing, mRNA degradation, translation, and protein degradation. The effects and propensity functions of these reactions are defined in Table \ref{tab:reaction_def}.

% see https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Regulation_of_gene_transcription
\begin{table}[h]
	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $w_G$, mRNA $x_G$ and proteins $y_G$ of gene $G$.} Define the set of regulators of $G$ as $R_G$, the set of upregulating regulators of $G$ as $R^+_G$, and the set of downregulating regulators of $G$ as $R^-_G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.} \label{tab:reaction_def}
	\centering
	\begin{tabular}{|lcc|}
		\hline
		Reaction & Effect & Propensity \\ \hline \hline
		Transcription & $\emptyset \rightarrow w_G$ & $wpr_G \times \frac{ba_G - 1 + \prod\limits_{H \in R^+_G}(1 + reg_{G,H})}{ba_G - 1 + \prod\limits_{H \in R_G}(1 + reg_{G,H})}$, \\
%		& & with $reg_{G,H} = \left(str_{G,H} \times \frac{y_H}{0.5 \times maxy_H}\right) ^ {co_{G,H}}$ \\
%		& & and $maxy_H = \frac{wpr_H \times ypr_H}{xdr_H \times ydr_H}$ \\
		Splicing & $w_G \rightarrow x_G$ & $wsr_G \times w_G$ \\
		mRNA degradation & $x_G \rightarrow \emptyset$ & $xdr_G \times x_G$ \\
		Translation & $x_G \rightarrow x_G + y_G$ & $ypr_G \times x_G$ \\
		Protein degradation & $y_G \rightarrow \emptyset$ & $ydr_G \times y_G$ \\ \hline
	\end{tabular}
\end{table}

% TODO: explain derivation of transcription formula? 
% regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]). y_H / maxy_H is more likely to be a probability that H is bound to G, though this does not take into account that H can be bound to other genes as well. 
% in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
% are independant and result in multiplicative effect.

% defining P_H is the probability that H is bound to G, 
% and the number of molecules produced C_H = ifelse(eff_G,H == 1, wpr_G × str_{G,H}, 0)
% should the expented no. of reactions (propensity) not be defined as:
% BA_G + Prod_H {C_H × P_H + (1 - P_H)} - Prod_H {1 - P_H} ?

@@

% max protein doesn't work as expected (given any simulation in model2):
%df <- data.frame(time = model2$simulations$meta$sim_time, as.matrix(model2$simulations$counts)) %>% 
%  gather(mol, value, -time) %>% 
%  filter(grepl("^y_", mol)) %>% 
%  mutate(feature_id = gsub("^y_", "", mol)) %>% 
%  left_join(model2$feature_info %>% transmute(feature_id, max_protein = wpr / xdr * ypr / ydr, module_id), by = "feature_id") %>% 
%  mutate(tfname = gsub("^[^_]*_", "", feature_id))
%ggplot(df) + geom_line(aes(time, value / max_protein, colour = tfname, group = feature_id)) + facet_wrap(~module_id)

\begin{table}[h]
	\caption{\textbf{Parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
	\centering
	\begin{tabular}{|lrl|}
		\hline
		Parameter & Symbol & Definition \\ \hline \hline
		Transcription rate & $wpr_G$ & $\in \{. | . \in N(100, 20), . \geq 10\}$ \\
		Splicing rate & $wsr_G$ & $\in \{. | . \in N(10, 2), . \geq 2\}$ \\
		mRNA degradation rate & $xdr_G$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
		Translation rate & $ypr_G$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
		Protein degradation rate & $ydr_G$ & $\in \{. | . \in N(3, 0.5), . \geq 1\}$ \\
		Interaction strength & $str_{G,H}$ & $\in \{10^. | . \in U(0, 2)\}$ * \\ 
		Interaction cooperativity & $co_{G,H}$ & $\in U(0.5, 2)$ * \\\hline\hline
		Max protein & $maxy_H$ & $= \frac{wpr_H \times ypr_H}{xdr_H \times ydr_H}$ \\ 
		Regulation activity & $reg_{G,H}$ & $= \left(str_{G,H} \times \frac{y_H}{0.5 \times maxy_H}\right) ^ {co_{G,H}}$ \\
		Basal expression & $ba_G$ & $= \begin{cases} 1 & \mbox{if } R^+_G = \emptyset \\ 0.0001 & \mbox{if } R^-_G = \emptyset \mbox{ and } R^+_G \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
		\multicolumn{3}{l}{*: unless already defined when $G$ is a TF.}
	\end{tabular}
\end{table}

>>>>

\newcommand{\w}[1]{\text{w}_{#1}}
\newcommand{\x}[1]{\text{x}_{#1}}
\newcommand{\y}[1]{\text{y}_{#1}}

\newcommand{\rs}[1]{\text{R}_{#1}}
\newcommand{\rp}[1]{\text{R}^+_{#1}}
\newcommand{\rn}[1]{\text{R}^-_{#1}}

\newcommand{\wpr}[1]{\text{wpr}_{#1}}
\newcommand{\wsr}[1]{\text{wsr}_{#1}}
\newcommand{\xdr}[1]{\text{xdr}_{#1}}
\newcommand{\ypr}[1]{\text{ypr}_{#1}}
\newcommand{\ydr}[1]{\text{ydr}_{#1}}

\newcommand{\str}[1]{\text{str}_{#1}}
\newcommand{\co}[1]{\text{co}_{#1}}
\newcommand{\ind}[1]{\text{ind}_{#1}}
\newcommand{\hmy}[1]{\text{hmy}_{#1}}
\newcommand{\reg}[1]{\text{reg}_{#1}}
\newcommand{\ba}[1]{\text{ba}_{#1}}

We define the abundance levels of pre-mRNA, mRNA and protein of gene $G$ as $\w G$, $\x G$ and $\y G$ respectively. Five reactions affect the abundance levels of these molecules: transcription, splicing, mRNA degradation, translation, and protein degradation. The effects and propensity functions of these reactions are defined in Table \ref{tab:reaction_def}.

% see https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Regulation_of_gene_transcription
\begin{table}[h]
	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.} \label{tab:reaction_def}
	\centering
	\begin{tabular}{|lcc|}
		\hline
		Reaction & Effect & Propensity \\ \hline \hline
		Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \reg{G,H})}{\prod\limits_{H \in \rp{G}}(1 + \reg{G,H})}$ \\
		Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
		mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
		Translation & $\x G \rightarrow \w G + \y G$ & $\ypr G \times \x G$ \\
		Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
	\end{tabular}
\end{table}

% TODO: explain derivation of transcription formula? 
% regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]). y_H / maxy_H is more likely to be a probability that H is bound to G, though this does not take into account that H can be bound to other genes as well. 
% in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
% are independent and result in multiplicative effect.

@@

\begin{table}[h]
	\caption{\textbf{Parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
	\centering
	\begin{tabular}{|lrl|}
		\hline
		Parameter & Symbol & Definition \\ \hline \hline
		Transcription rate & $\wpr{G}$ & $\in \{. | . \in N(100, 20), . \geq 10\}$ \\
		Splicing rate & $\wsr G$ & $\in \{. | . \in N(10, 2), . \geq 2\}$ \\
		mRNA degradation rate & $\xdr{G}$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
		Translation rate & $\ypr{G}$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
		Protein degradation rate & $\ydr G$ & $\in \{. | . \in N(3, 0.5), . \geq 1\}$ \\
		Interaction strength & $\str{G,H}$ & $\in \{10^. | . \in U(0, 2)\}$ * \\ 
		Interaction cooperativity & $\co{G,H}$ & $\in U(0.5, 2)$ * \\
    Independence factor & $\ind G$ & $\in [0, 1]$ \\ \hline\hline
		TF concentration at half-maximal binding & $\hmy H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\ 
		Regulation activity & $\reg{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\hmy H}\right) ^ {\co{G,H}}$ \\
		Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
		\multicolumn{3}{l}{*: unless already defined when $G$ is a TF.}
	\end{tabular}
\end{table}
